---
title: "Network Physiology with PhysioNetPhysiology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network Physiology with PhysioNetPhysiology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
set.seed(1)
```

```{r setup}
library(PhysioNetPhysiology)
```

## What this package does

*Network Physiology* studies how distinct physiological systems — brain,
heart, respiration, muscle — **dynamically interact**. A single-modality tool
cannot ask that question; this package can, because it works over a node matrix
in which each column is one physiological system.

The engine is **Time Delay Stability (TDS)** (Bashan et al. 2012): slide a window
along the recording, and in each window find the lag that maximises a pair's
cross-correlation. Where that lag stays put for a stretch of time, the two
systems are stably coupled. Stable couplings become links in an
*organ-interaction network*.

## A node matrix

Nodes are physiological features sampled on a common (usually slow) grid.
Here we simulate four systems where `brain` drives `heart` (a fixed lag), and
`heart` drives `resp` only during the second half of the recording; `muscle`
is independent.

```{r nodes}
n <- 4000; half <- n / 2
drive <- function() { s <- as.numeric(stats::filter(rnorm(n), rep(1/8, 8), sides = 2)); s[is.na(s)] <- 0; s }
brain  <- drive()
heart  <- c(rep(0, 4), brain[seq_len(n - 4)]) + rnorm(n, sd = 0.4)   # brain -> heart (lag 4)
resp   <- rnorm(n, sd = 0.4)
resp[(half + 4):n] <- heart[half:(n - 3)] + rnorm(half - 3, sd = 0.4) # heart -> resp, 2nd half
muscle <- rnorm(n)

X <- physioNodeMatrix(list(brain = brain, heart = heart,
                           resp = resp, muscle = muscle))
```

## Time Delay Stability

```{r tds}
tds <- timeDelayStability(X, window = 200, step = 100, max_lag = 30, min_stable = 4)
summary(tds)
```

`brain–heart` is stable across the whole recording; `heart–resp` and the
indirect `brain–resp` appear because respiration is coupled for half the time.

## Organ-interaction network

Links are kept only if they clear a significance threshold estimated from
phase-randomised surrogates (which preserve each signal's own spectrum but
destroy cross-signal timing).

```{r network}
net <- tdsNetwork(tds, surrogate = TRUE, X = X, n_surrogates = 30)
net
plotTDSnetwork(net)
```

## Network reconfiguration across states

The signature Network-Physiology phenomenon: the network reorganises across
physiological states. Here the state switches at the midpoint.

```{r state}
state <- rep(c("A", "B"), each = half)
bystate <- tdsNetworkByState(X, state, window = 200, step = 100,
                             max_lag = 30, min_stable = 4, threshold = net$threshold)
bystate$reconfiguration
tdsReconfiguration(bystate, from = "A", to = "B")
```

The `heart–resp` (and indirect `brain–resp`) couplings appear only in state B.

## Cardiorespiratory coupling

For the heart–respiration relationship specifically, phase-based tools quantify
n:m locking. Here we simulate four heartbeats per breath.

```{r cardioresp}
fs <- 100; t <- seq(0, 120, by = 1 / fs)
resp_sig <- sin(2 * pi * 0.25 * t)
card_sig <- sin(2 * pi * 1.00 * t)
rpeaks   <- which(diff(sign(card_sig)) > 0)

cr <- cardiorespiratoryCoupling(resp_sig, rpeaks, fs)
cr
```

## Brain–heart: the heartbeat-evoked potential

```{r hep}
n2 <- 60 * 250; fs2 <- 250
rp <- seq(fs2, n2 - fs2, by = round(0.9 * fs2))
eeg <- rnorm(n2)
bt <- round(0.3 * fs2); w <- round(0.04 * fs2)
for (r in rp) { idx <- (r + bt - 3*w):(r + bt + 3*w); idx <- idx[idx >= 1 & idx <= n2]
  eeg[idx] <- eeg[idx] + 3 * exp(-((idx - (r + bt))^2) / (2 * w^2)) }

hep <- heartbeatEvokedPotential(eeg, rp, fs2)
hep
plotHEP(hep)
```

## Working with PhysioExperiment objects

`physioNodeMatrix()` and `timeDelayStability()` also accept a `PhysioExperiment`
(each channel becomes a node) or a `MultiPhysioExperiment` (each experiment
contributes nodes), extracting a `"raw"` or `"envelope"` feature and optionally
down-sampling to a common rate — so a multimodal recording flows straight into
the TDS engine.

## References

- Bashan A, Bartsch RP, Kantelhardt JW, Havlin S, Ivanov PC (2012). Network
  physiology reveals relations between network topology and physiological
  function. *Nat Commun* 3:702.
- Bartsch RP, Liu KKL, Bashan A, Ivanov PC (2015). Network physiology: how organ
  systems dynamically interact. *PLoS ONE* 10(11):e0142143.
